Exploiting delayed dichromatism to disentangle the effects of adult survival and recruitment on the population dynamics of waterfowl

### 2022-10-19 Code/Draft_ruddy_duck_vital_rates.Rmd
graphs for article

Assessing population growth-rate is a key step towards a better understanding of factors underlying the dynamics of natural populations [e.g. @Niel2005]. It is also, crucial for measuring the efficacy of management actions eventually undertaken to help population recovery, control or eradication [@Shea1998]. Among the approaches available to managers for reaching these goals, those relying on the monitoring of the numbers of individuals are by far the most popular [@Rintala2022]. In many instances, these methods even enable environmental factors underlying changes in population size to be investigated and hence some mitigating actions to be implemented [@Faillettaz2019]. Unfortunately, relying on the monitoring of numbers of individuals alone hinder basic understanding of the demographic mechanisms underlying changes in population growth rates. More specifically, based on counts only, it is nearly impossible to assess to what extent population growth rate is affected by change survival rather than to variations in reproductive parameters [@Austin2000]. To assess the relative sensitivity of population growth rate to factors affecting survival or reproductive parameters, demographers most often rely on the monitoring of individuals by capture-mark-recapture [@Lebreton1992]. In practice however, one cannot always rely on the capture and release of individuals owing to the legal status of the species that preclude any additional disturbance (critically endangered species), or forbid the release of alive individuals (invasive species or pests). Indeed, although highly efficient for assessing demographic parameters, CMR methods have potential drawbacks such as being invasive and hardly affordable when time and money are limiting. Genetic monitoring is a non-invasive alternative to CMR approaches based on physical captures, but it suffers from being costly and time-consuming and requires quite rigorous sampling schemes in the field.

Most of the time therefore, managers do the best of a bad job using counts as the only viable option for assessing population growth rates and the efficacy of management actions. Counting the number of immatures produced in addition to alive adults may enable adult survival and recruitment rate to be estimated, and hence provides knowledge on which of adult survival and productivity has the most influence on population growth rate. Unfortunately, this is only possible for a limited number of species in which broodless adults and those with young display the same detectability. Alternatively, assessing the sex and age structure in hunting bags has been used to infer the role of decreasing reproductive success in population declines in a number of games species including ducks and geese [@Fox2018]. However, this approach is not suited for protected/endangered species. Here we exemplify how, in dimorphic species, delayed maturity of males can be exploited for estimating productivity/recruitment and adult survival by differentiating male-like to female-like individuals in repeated winter counts. We use two non-native Ruddy duck populations introduced in Europe as a study model. As numerous duck species, the Ruddy duck is dimorphic, with newly born males looking like females until the pre-nuptial moult, which occurs in late winter. This means that the apparent proportion of males increases over the course of the wintering season. These changes in apparent proportions of males during this period are therefore directly related to the proportion of immatures into the populations and hence to the reproductive success and the recruitment rate. This study details the method to infer adult survival and recruitment rate from count data and an application is developed to assess the impact of two different eradication strategies deployed in Great Britain and in France respectively.

@@

The ruddy duck is a stiff-tailed duck native from the Americas. From seven individuals initially acclimated in the Slimbridge Wetland Centre in the Great Britain in the 40’s [@GutierrezExposito2020], a feral population began to establish with the first observed reproductive attempts in the wild in the 60s. This feral population rapidly grew and spread to the entire country to reach more than 5000 individuals in the early 2000s. By the end of the 80s, a feral population began to establish also on the continent in France, supposedly owing to the arrival of individuals born in Great Britain (GB). Contrary to what was observed in GB, the French Ruddy duck population did not spread much, with the vast majority of observations and breeding attempts clustering in Western France. In France, during winter, almost none Ruddy duck is observed outside the lake of Grand Lieu (47.09°N, 1.67°W), which is thus the exclusive wintering site and greatly facilitates the monitoring of this population.

The feral European populations of Ruddy duck are considered as a major threat to the endangered native White-headed duck population of the South-Western Mediterranean because of hybridization risks and thus an elevated potential for genetic pollution and eventually genetic assimilation of the latter by the former [@MunozFuentes2007]. In order to mitigate the risks of genetic pollution of the White-headed duck by Ruddy ducks a European plan of eradication of Ruddy ducks has been adopted by the European Commission. As result, eradication measures were taken both in France and Great Britain in the late 90s [@GutierrezExposito2020].

The release non-native species in the wild, including Ruddy ducks, is forbidden in countries of the European Union, and the status of White-headed duck populations is highly unfavourable. The use of capture-mark-recapture to monitor populations of these species was thus impossible. As a result, the effectiveness of the eradication of Ruddy ducks and management actions intended to favour the  recovery of White-headed ducks were essentially checked through censuses. Censuses, however, prevented from assessing the relative effects or variations in survival and reproductive success/recruitment on population growth rate, which is a key towards a proper understanding of the efficacy of management actions.

Like many other ducks, Ruddy ducks and White-headed ducks both display delayed dichromatism. Thereby immature males acquire the typical colourful plumage of their kind over the course of the interbreeding season, which is typically in late winter. Immature males are thus confounded to immature and adult females before their moult (Figure \@ref(fig:flock)). The delayed dichromatism explains why the apparent proportion of females in the winter counts is always higher than the true proportion of females that can be observed in winter hunting bags (Figure \@ref(fig:pfem)). Thus, by monitoring the seasonal evolution of apparent sex-ratio, it is possible to assess the proportion of immature males within a population, which, combined with reliable counts and under some reasonable assumptions (e.g., even sex-ratio at birth, comparable mortality rates between females and males among juveniles, and unchanging adult sex-ratio over the monitoring period), enables both adult survival rate and recruitment rate to be estimated.

@@

We monitored the demography of Ruddy duck populations in both Great-Britain and France thanks to exhaustive counts on the wintering grounds during the period between December 1 and January 31. In addition to these exhaustive counts, censuses that distinguished female-like from male-like individuals were performed on the same period from 2007 to 2012 in UK, and in 1999, 2001-2009, 2012-2019 in France. During these years, the eradication effort, under control of the authorities in both countries, was also scrupulously recorded.

@@

|Name|Class|Description|
|:---|:----|:----------------------------|
|$AM$|Data|Total number of adult males in hunting bags|
|$AF$|Data|Total number of adult females in hunting bags|
|$p(m|a)$|Parameter|Proportion of males in adults (or probability to be a male knowing it is an adult)|
|$CML_{i, t}$|Data|Cumulated number of male-like individuals counted in population $i$ in year $t$, which are assumed to be adult males|
|$C_{i, t}$|Data|Cumulated number of individuals counted in population $i$ in year $t$|
|$p(a \cap m)_{i, t}$|Parameter|Proportion of adult males in population $i$ in year $t$|
|$p(a)_{i, t}$|Parameter|Proportion of adults in population $i$ in year $t$|
|$p(i)_{i, t}$|Parameter|Proportion of immatures in population $i$ in year $t$|
|$N_{i, t}$|Data|Size of the population $i$ in year $t$ (maximum number of individuals counted in the wintering population)|
|$s_{i, t}$|Parameter|Adult survival rate, i.e. proportion of adults in year $t-1$ still alive in year $t$|
|$r_{i, t}$|Parameter|Recruitment rate, i.e. number of recruits in population $i$ in year $t$ per adult in year $t - 1$|
|$S_{i, t}$|Parameter|Number of adults in population $i$ in year $t$|
|$R_{i, t}$|Parameter|Number of recruits in population $i$ in year $t$|

@@

\begin{equation}
s_{i, t} = \frac{(1 - p(i)_{i, t}).N_{i, t}}{N_{i, t - 1}}
r_{i, t} = \frac{p(i)_{i, t}.N_{i, t}}{N_{i, t - 1}}
(\#eq:vr)
\end{equation}

\begin{equation}
S_{i, t} = (1 - p(i)_{i, t}).N_{i, t}
R_{i, t} = p(i)_{i, t}.N_{i, t}
(\#eq:VR)
\end{equation}

## Validation of the count method {#VAL}

The relevance of the method was assessed by testing the likelihood of the estimates of vital rates. For the adult survival rate, we checked if the values belonged to the expected interval $[0;1]$, and we compared the average values to literature data, since the adult survival is commonly assessed by capture mark recapture [@Lebreton2001]. For the recruitment rate, which is rarely explicitly informed in the literature, we checked if the values were equal or below the maximum recruitment rate expected without exploitation and without negative density-dependence processes. If the recruitment rate outputs of the model are significantly superior to the maximum recruitment rate, it means that the method overestimates the recruitment rate, which reflects that a part of adult males are not detected during count surveys. A null or negative recruitment rate would mean that immature males moulted before the survey or/and the assumption on the stability of the adult sex ratio is too strong.

The maximum recruitment rate is deduced by using its relationship to the maximum growth rate and the maximum survival, see Equation \@ref(eq:GR). This relationship becomes more complex if a species with delayed maturity is considered, see @Robertson2008. This relationship comes from a simple reasoning for a closed population: the population size in year $t$ is equal to the number of adults that survived over year $t - 1$ plus the offspring produced in year $t - 1$ that survived until the reproduction period of year $t$, i.e. the recruitment in year $t$. The growth rate of a population is thus the sum of the adult survival rate and the recruitment rate [@Flint2015]. For an open population, adult survival and recruitment rate are confounded to adult and recruit migration respectively.

|Name|Description|
|:--|:----------------------------|
|$N_{t}$|Number of adults in year $t$|
|$D_{t}$|Number of adults dead during year $t$|
|$R_{t}$|Number of recruits in year $t$|
|$s_{t}$|Adult survival rate, i.e. proportion of adults in year $t-1$ still alive in year $t$|
|$r_{t}$|Recruitment rate, i.e. number of recruits in year $t$ produced per adult in year $t-1$|
|$\lambda_{t}$|Growth rate of the population between year $t-1$ and year $t$|

@@

The maximum growth rate occurred for both populations during their expansion phase, when the evolution of the population size is the steepest. To estimate robust maximum population growth rates for both populations, we smoothed annual population growth rate over a consistent time period by using a linear regression on the logarithm scale, see Equation \@ref(eq:L).

@@

Practically, we used a very conservative prior on the maximum adult survival by setting a uniform distribution on $[0.7; 0.9]$ [the lower limit corresponds to the upper range of survival rates observed in literature for species of similar weight, the upper limit corresponds to the range of survival rates of long-life species, @Nichols1997; @Krementz1997; @Buxton2004]. For the maximum growth rate, we rejected GB data before 1972 since the size estimate of this population is noisy below 50 individuals (Figure \@ref(fig:grts)). After reaching the threshold of 1000 individuals, the GB population growth observed a strong inflexion whereas no harvest pressure was applied. This observation suggests that a negative density-dependence process might occur beyond 1000 individuals and led to consider only the first sequence of the time series to infer on the maximum growth rate in GB. For the French population, the two sequences with birds and without harvest were used, i.e. 1994-1999 and 2001-2004 (Figure \@ref(fig:grts)).

```{r grts, fig.cap = "(ref:grts)"}
knitr::include_graphics("../Output/plot_pres_1.png")
```

Even if both vital rates vary in a likely range following the validation methodology previously described, it does not prove that the interannual variability is well tracked. In order to validate this aspect, we compared the outputs of the count-based method, i.e. the immature proportion and both vital rates, to outputs based on data from hunting bags. Only data from years with more than 100 individuals harvested in winter were selected, which corresponded to a time series covering 5 years of the GB count time series. The presence of the bursa of Fabricius enables immature to be identified [@Hochbaum1942]. The immature proportion in hunting bags is then straightforward, see Equation \@ref(eq:pis). The adult survival and the recruitment rate are then estimated by combining this proportion and Equation \@ref(eq:vr). We discuss the ability of the count method to track the interannual variability by comparing the outputs of the two methods.

@@

Harvest strategies differ from GB to France. In GB, harvest occurred in winter, i.e. before the breeding period, whereas it is done mostly in summer in France, i.e. during and after the breeding period. To assess the impact of the two strategies on the population dynamics and conclude on which strategy is the most suited to reach the management objectives, we compared vital rates with and without harvest in light of the underlying population growth rates.

For all the sub-models in Section \@ref(VAL), the Bayesian framework was used for its efficiency and simplicity to propagate error through the parameters. We used uninformative priors on all parameters. We generated three chains of length 500000, with a thinning of 10 to avoid autocorrelation in the samples, and we discarded the first 2000 samples as burn-in. Chain convergence was assessed using the Gelman and Rubin convergence diagnostic (R<1.1, @Gelman1992). We fit the models using NIMBLE [@Valpine2017] run from R [@RCT2022]. Data and code are available here: https://github.com/adri-tab/Ruddy_duck_vital_rates. The values **X[Y; Z]** reported in results are the medians and the associated boundaries of the 95% confidence interval of posterior distributions. The median was preferred to the mean because of its robustness to skewed distribution.

@@

Time series derived from these counts show that the two populations display comparable growth (Figure \@ref(fig:count)). They grew freely in both countries until 1999. Although shooting was initiated in 1999 in both countries, both populations begun declining only after much higher harvest efforts had been achieved in 2005 and 2018 respectively (figure 3). This history underlines that both populations observed a large spectrum of harvest effort, from no pressure to very high pressure. For sake of interpretation, we ascribed the time-period 2001-2004 to the no harvest category in France because the harvest pressure was negligible hand hence had no effect on the population growth.

@@

```{r count, fig.cap = "(ref:count)"}
knitr::include_graphics("../Output/plot_2.png")
```

@@

(ref:flock) Typical observation of a ruddy duck flock in winter: 10 female-like individuals with a whitish cheek and a dark stripe across it, 4 male-like individuals with a white cheek and a black cap, 3 unidentified individuals © Jay McGowan - 3 February 2013 - Tompkins, New York, United States

(ref:pfem) Apparent female proportion in winter counts vs female proportion in hunting bags; for the first category, one point corresponds to the proportion of female-like plumage in a population counted in winter; for the second category, the female proportion is estimated from all ruddy ducks controlled over one year in a single population

(ref:count) Evolution of the size of the two ruddy duck populations in light of the harvest pressure and the data time series

(ref:grts) Evolution of the size of the two ruddy duck populations in light of the harvest pressure; on time periods without harvest, maximum population growth rates can be estimated (slopes in red for the UK and in blue for France)

>>>>

Assessing population growth-rate is a key step towards a better understanding of factors underlying the dynamics of natural populations [e.g. @Niel2005]. It is also crucial for measuring the efficacy of management actions eventually undertaken to help population recovery, control, or eradication [@Shea1998]. Among the approaches available to managers for reaching these goals, those relying on the monitoring of the number of individuals are by far the most popular [@Rintala2022]. In many instances, these methods even enable environmental factors underlying changes in population size to be investigated and hence some mitigating actions to be implemented [@Faillettaz2019]. Unfortunately, relying on the monitoring of numbers of individuals alone hinder basic understanding of the demographic mechanisms underlying changes in population growth rates. More specifically, based on counts only, it is nearly impossible to assess to what extent population growth rate is affected by change in survival rather than to variations in reproductive parameters [@Austin2000]. To assess the relative sensitivity of population growth rate to factors affecting survival or reproductive parameters, demographers most often rely on the monitoring of individuals by capture-mark-recapture [@Lebreton1992]. In practice however, one cannot always rely on the capture and release of individuals owing to the legal status of the species that precludes any additional disturbance, e.g. critically endangered species, or forbid the release of alive individuals, e.g. invasive species or pests. Indeed, although highly efficient for assessing demographic parameters, CMR methods have potential drawbacks such as being invasive and hardly affordable when time and money are limiting. Genetic monitoring is a non-invasive alternative to CMR approaches based on physical captures, but it suffers from being costly and time-consuming and requires quite rigorous sampling schemes in the field.

Most of the time therefore, managers do the best of a bad job using counts as the only viable option for tracking population growth rates and assessing the efficiency of management actions. Counting the number of immatures produced in addition to alive adults may enable adult survival and recruitment rate to be estimated, and hence provides knowledge on which of adult survival and productivity has the most influence on population growth rate. Unfortunately, this is only possible for a limited number of species in which broodless adults and those with young display the same detectability. Alternatively, assessing the sex and age structure in hunting bags has been used to infer the role of decreasing reproductive success in population declines in a number of games species including ducks and geese [@Fox2018]. However, this approach is not suited for protected/endangered species. Here we exemplify how, in dimorphic species, delayed maturity of males can be exploited for estimating productivity/recruitment and adult survival by differentiating male-like to female-like individuals in repeated winter counts. We use two non-native Ruddy duck populations introduced in Europe as a study model. As numerous duck species, the Ruddy duck is dimorphic, with newly born males looking like females until the pre-nuptial moult, which occurs in late winter. This means that the apparent proportion of males increases over the course of the wintering season. These changes in apparent proportions of males during this period are therefore directly related to the proportion of immatures into the populations and hence to the reproductive success and the recruitment rate. This study details the method to infer adult survival and recruitment rate from count data and an application is developed to assess the impact of two different eradication strategies deployed in Great Britain and in France respectively.

@@

The ruddy duck is a stiff-tailed duck native from the Americas. From seven individuals initially acclimated in the Slimbridge Wetland Centre in the Great Britain in the 40’s [@GutierrezExposito2020], a feral population began to establish with the first observed reproductive attempts in the wild in the 60s (Figure \@ref(fig:count)). This feral population rapidly grew and spread to the entire country to reach more than 5000 individuals in the early 2000s. By the end of the 80s, a feral population began to establish also on the continent in France, supposedly owing to the arrival of individuals born in Great Britain. Contrary to what was observed in Great Britain, the French Ruddy duck population did not spread much, with the vast majority of observations and breeding attempts clustering in Western France. In France, during winter, almost none Ruddy duck is observed outside the lake of Grand Lieu (47.09°N, 1.67°W), which is thus the exclusive wintering site and greatly facilitates the monitoring of this population. The feral European populations of Ruddy duck are considered as a major threat to the endangered native White-headed duck population of the South-Western Mediterranean because of hybridization risks and thus an elevated potential for genetic pollution and eventually genetic assimilation of the latter by the former [@MunozFuentes2007]. In order to mitigate the risks of genetic pollution of the White-headed duck by Ruddy ducks, a European plan of eradication of Ruddy ducks has been adopted by the European Commission. As a result, eradication measures were taken both in Great Britain and in France from the late 90s [@GutierrezExposito2020], (Figure \@ref(fig:count)). The release of non-native species in the wild, including Ruddy ducks, is forbidden in countries of the European Union, and the status of White-headed duck populations is highly unfavourable. The use of capture-mark-recapture to monitor populations of these species is thus impossible. As a result, the effectiveness of the eradication of Ruddy ducks and management actions intended to favour the recovery of White-headed ducks were essentially checked through censuses. Censuses, however, prevented from assessing the relative effects or variations in survival and reproductive success/recruitment on population growth rate, which is a key towards a proper understanding of the efficiency of management actions.

Like many other ducks, Ruddy ducks and White-headed ducks both display delayed dichromatism. Thereby immature males acquire the typical colourful plumage of their kind over the course of the interbreeding season, which is typically in late winter. Immature males are thus confounded to immature and adult females before their moult (Figure \@ref(fig:flock)). The delayed dichromatism explains why the apparent proportion of females in the winter counts is always higher than the true proportion of females that can be observed in winter hunting bags (Figure \@ref(fig:pfem)). Thus, it is possible to assess the proportion of immatures within a population by monitoring the seasonal evolution of apparent sex ratio and assuming an even sex ratio at birth, comparable mortality rates between females and males among juveniles, and unchanging adult sex ratio over the monitoring period. Then, both adult survival rate and recruitment rate can be estimated by combining the age ratio thus obtained with reliable counts over two successive winters.

@@

We monitored the demography of Ruddy duck populations in both Great-Britain and France thanks to exhaustive counts on the wintering grounds during the period between December 1 and January 31. In addition to these exhaustive counts, censuses that distinguished female-like from male-like individuals were performed on the same period from 2007 to 2012 in Great Britain, which corresponded to a strong decline of the population. In France, the monitoring spread over 1999, 2001-2009, and 2012-2019, which corresponded to a stabilized population size excepted over 2001-2004, for which the population grew rapidly without control effort. During these years, the eradication effort, under control of the authorities in both countries, was also scrupulously recorded.

@@

|Name|Class|Description|
|:---|:----|:----------------------------|
|$AM$|Data|Total number of adult males in hunting bags|
|$AF$|Data|Total number of adult females in hunting bags|
|$p(m|a)$|Parameter|Proportion of males in adults (or probability to be a male knowing it is an adult)|
|$CML_{i, t}$|Data|Cumulated number of male-like individuals counted in population $i$ in year $t$, which are assumed to be adult males|
|$C_{i, t}$|Data|Cumulated number of individuals counted in population $i$ in year $t$|
|$p(a \cap m)_{i, t}$|Parameter|Proportion of adult males in population $i$ in year $t$|
|$p(a)_{i, t}$|Parameter|Proportion of adults in population $i$ in year $t$|
|$p(i)_{i, t}$|Parameter|Proportion of immatures in population $i$ in year $t$|
|$N_{i, t}$|Data|Size of the population $i$ in year $t$ (maximum number of individuals counted in the wintering population)|
|$s_{i, t}$|Parameter|Adult survival rate, i.e. proportion of individuals in year $t-1$ still alive in year $t$|
|$r_{i, t}$|Parameter|Recruitment rate, i.e. number of recruits in population $i$ in year $t$ per individual in year $t - 1$|
|$S_{i, t}$|Parameter|Number of adults in population $i$ in year $t$|
|$R_{i, t}$|Parameter|Number of recruits in population $i$ in year $t$|

@@

\begin{equation}
\begin{aligned}
s_{i, t} & = \frac{N_{i, t}}{N_{i, t - 1}}.(1 - p(i)_{i, t}) \\
r_{i, t} & = \frac{N_{i, t}}{N_{i, t - 1}}.p(i)_{i, t}
\end{aligned}
(\#eq:vr)
\end{equation}

\begin{equation}
\begin{aligned}
S_{i, t} & = N_{i, t}.(1 - p(i)_{i, t}) \\
R_{i, t} & = N_{i, t}.p(i)_{i, t}
\end{aligned}
(\#eq:VR)
\end{equation}

## Validation of the count-based method {#VAL}

The relevance of the method was assessed by testing the likelihood of the estimates of vital rates. For the adult survival rate, we checked if the values belonged to the expected interval $[0;1]$, and we compared the average values to literature data, since the adult survival is commonly assessed by capture mark recapture [@Lebreton2001]. For the recruitment rate, which is rarely explicitly informed in the literature, we checked if the values were equal or below the maximum recruitment rate expected without exploitation and without negative density-dependence processes. If the recruitment rate outputs of the model are significantly superior to the maximum recruitment rate, it means that the method overestimates the recruitment rate, which reflects that a part of adult males are not detected during count surveys. A null or negative recruitment rate would mean that immature males moulted before the survey or/and the assumption on the stability of the adult sex ratio is overrated.

The maximum recruitment rate is deduced by using its relationship to the maximum growth rate and the maximum survival, see Equation \@ref(eq:GR). This relationship comes from a simple reasoning for a closed population: the population size in year $t$ is equal to the number of adults that survived over year $t - 1$ plus the offspring produced in year $t - 1$ that survived until the reproduction period of year $t$, i.e. the recruitment in year $t$. The growth rate of a population is thus the sum of the adult survival rate and the recruitment rate [@Flint2015]. For an open population, adult survival and recruitment rate are confounded to adult and recruit migration respectively. This relationship becomes more complex if a species with delayed maturity is considered, see @Robertson2008.

|Name|Description|
|:--|:----------------------------|
|$N_{t}$|Number of adults in year $t$|
|$D_{t}$|Number of adults dead during year $t$|
|$R_{t}$|Number of recruits in year $t$|
|$s_{t}$|Adult survival rate, i.e. proportion of individuals in year $t-1$ still alive in year $t$|
|$r_{t}$|Recruitment rate, i.e. number of recruits in year $t$ produced per individual in year $t-1$|
|$\lambda_{t}$|Growth rate of the population between year $t-1$ and year $t$|

@@

The maximum growth rate occurred for both populations during their expansion phase, when the evolution of the population size was the steepest. To estimate robust maximum population growth rates for both populations, we smoothed annual population growth rate over a consistent time period by using a linear regression on the logarithm scale, see Equation \@ref(eq:L).

@@

Practically, we used a very conservative prior on the maximum adult survival by setting a uniform distribution on $[0.7; 0.9]$. The lower limit corresponds to the upper range of survival rates observed in literature for species of similar weight, the upper limit corresponds to the range of survival rates of long-life species [@Nichols1997; @Krementz1997; @Buxton2004]. For the maximum growth rate, we rejected Great Britain data before 1972 since the size estimate of this population is noisy below 50 individuals (Figure \@ref(fig:count)). After reaching the threshold of 1000 individuals, the Great Britain population growth observed a strong inflexion whereas no harvest pressure was applied. This observation suggests that a negative density-dependence process might occur beyond 1000 individuals and led to consider only the first sequence of the time series to infer on the maximum growth rate in Great Britain. For the French population, the two sequences without harvest were used, i.e. 1994-1999 and 2001-2004.

Even if both vital rates vary in a likely range following the validation methodology previously described, it does not prove that the interannual variability is properly tracked. In order to validate this aspect, we compared the outputs of the count-based method, i.e. the immature proportion and both vital rates, to outputs based on data from hunting bags. Only data from years with more than 100 individuals harvested in winter were selected. It covers 5 years of the count time series of the Great Britain population. The presence of the bursa of Fabricius enables immature to be identified [@Hochbaum1942]. The estimation of the immature proportion in hunting bags is then straightforward, see Equation \@ref(eq:pis). The adult survival and the recruitment rate were then estimated by combining this proportion and Equation \@ref(eq:vr). We discussed the ability of the count-based method to accurately track the interannual variability by comparing the outputs of the two methods.

@@

Harvest strategies differed from Great Britain to France. In Great Britain, harvest occurred in winter, i.e. before the breeding period, whereas it was done mostly in summer in France, i.e. during and after the breeding period. To identify the impact of the two strategies on the population dynamics, we assessed how vital rates responded to the harvest by considering both average drops and variability changes. To complete our understanding on why the Great Britain strategy was more performing on controlling Ruddy ducks, we compared vital rates with and without harvest in light of the underlying average population growth rates.

For all the sub-models in Section \@ref(MM), the Bayesian framework was used for its efficiency and simplicity to propagate error through the parameters. We used uninformative priors on all parameters. We generated three chains of length 500000, with a thinning of 10 to avoid autocorrelation in the samples, and we discarded the first 2000 samples as burn-in. Chain convergence was assessed using the Gelman and Rubin convergence diagnostic (R<1.1, @Gelman1992). We fit the models using NIMBLE [@Valpine2017] run from R [@RCT2022]. Data and code are available here: https://github.com/adri-tab/Ruddy_duck_vital_rates. The values **X[Y; Z]** reported in results are the medians and the associated boundaries of the 95% confidence interval of posterior distributions. The median was preferred to the mean because of its robustness to skewed distribution.

@@

Higher mortality on adult in GB -> better detection in winter;

Time series derived from count data show that the two populations display comparable initial growth (Figure \@ref(fig:count2)). They grew freely in both countries until 1999. Although shooting was initiated in 1999 in both countries, both populations begun declining only after much higher harvest efforts had been achieved in 2005 and 2018 respectively (figure 3). This history underlines that both populations observed a large spectrum of harvest effort, from no pressure to very high pressure. For sake of interpretation, we ascribed the time-period 2001-2004 to the no harvest category in France because the harvest pressure was negligible hand hence had no effect on the population growth.

@@

@@

(ref:count) Evolution of the size of the two Ruddy duck populations; y-axis is on a log-scale

(ref:flock) Typical observation of a Ruddy duck flock in winter: 10 female-like individuals with a whitish cheek and a dark stripe across it, 4 male-like individuals with a white cheek and a black cap, 3 unidentified individuals © Jay McGowan - 3 February 2013 - Tompkins, New York, United States

(ref:pfem) Female proportion in hunting bags vs in counts; For the first category, one point corresponds to the female proportion estimated from direct sex identification of the Ruddy ducks controlled over one year in a population; for the second category, one point is the proportion of female-like plumage in a population counted in winter. The difference is due to immature males that look like females before moulting

(ref:grts) Evolution of the size of the two Ruddy duck populations in light of the harvest pressure; on time periods without harvest, maximum population growth rates can be estimated (slopes in red for the UK and in blue for France)
